Many claims of group polarization are plausibly false due to inappropriate statistics

### 2022-11-30 GroupPolarizationStatmod/paper/ms.tex
paper updates to model section especially

% HEADINGS
\usepackage{sectsty} 
\usepackage[normalem]{ulem} 
\sectionfont{\upshape\Large}
\subsectionfont{\scshape\normalsize} 
\subsubsectionfont{\upshape\large}

@@

\newcommand{\sigmapre}{\sigma_\mathrm{pre}}
\newcommand{\sigmapost}{\sigma_\mathrm{post}}
\newcommand{\normalpre}{\mathcal{N}(\mu, \sigmapre)}
\newcommand{\normalpost}{\mathcal{N}(\mu, \sigmapost)}

@@

\begin{table}[h]
  \begin{tabular}{cp{5.0in}} \toprule
   Symbol & Description  \\ \midrule  
   $t \in \{\mathrm{pre}, \mathrm{post}\}$ & Time, either \emph{pre}- or
                                           \emph{post}-discussion \\
   $k \in \{0,\ldots,K\} $ & Bin index ($k$) between 0 and total number of
         ordinal bins, $K$ \\
   $\mu$  &   Latent mean, assumed constant in simple consensus \\ 
   $p(o;\mu,\sigma_t)$ & Normal latent distribution of opinions with mean
                         $\mu$ and variance $\sigma^2$ \\
   $p(o \in b_k; \mu, \sigma_t)$ & Ordinal density of binned opinions in bin $b_k$ 
            (Equation~\ref{eq:binFrequency}) \\
   $\theta_k$  & The upper threshold value for binning latent opinions into
   bin $b_k$ \\
   $\langle o_t \rangle$  &   Observed mean opinion at time $t$ \\
   \bottomrule
  \end{tabular} 
\end{table}

@@

@@

To evaluate whether a positive group polarization observation is plausibly false,
we develop two layers of evidence for plausible falsehood. The first layer
tests whether we can identify $\sigmapre$ and

We used large-$N$ and generative modeling supported by $t$-tests on simulated
data to demonstrate that many published findings are plausibly false.

We model measurement of $N \to \infty$ participant opinions,
which results in a histogram of frequency of opinions
in each bin, i.e., $o_{i,t} = b_k$. The frequency of responses in each bin is 
the integral over the continuous normal probability density function from 
one bin threshold to another. This transforms the probability density 
function from $p(o_{i,t};~\mu_t, \sigma_t)$ to the probability
of observing a reported opinion of each bin value. 
The probability of observing an opinion in bin
$b_k$ is calculated by integrating the normal probability density 
function over the range of $\theta_{k-1}$ to $\theta_{k}$. 
Formally, we write the 
probability of observing an opinion in bin $k$ at time $t$ as
\begin{equation}
\begin{aligned}
  p(o \in b_k; \mu_t, \sigma_t, \theta) 
    &= \int_{\theta_{k-1}}^{\theta_k} p(o;\mu_t, \sigma_t) do \\
    &= \Phi \left( \frac{o - \mu_t}{\sigma_t} \right)\Big|_{o=\theta_{k-1}}^{\theta_k} \\
    &= \Phi \left( \frac{\theta_k - \mu_t}{\sigma_t} \right) - 
       \Phi \left( \frac{\theta_{k-1} - \mu_t}{\sigma_t} \right)
\end{aligned}
\label{eq:binFrequency}
\end{equation}
\noindent
where $\Phi(\frac{o - \mu}{\sigma})$ is the normalized 
normal cumulative distribution function over opinions $o$, shifted by an amount $\mu$ with
standard deviation $\sigma$.
\subsubsection{Identify null-model parameters that yield plausibly-false positives}

@@

After identifying null-model parameters that yield plausibly-false positives 
in the $N \to \infty$ case, we generated binned samples from normal opinion 
distributions $\normalpre$ and $\normalpost$.

@@

>>>>

% HEADINGS
% \usepackage{sectsty} 
% \usepackage[normalem]{ulem} 
% \sectionfont{\upshape\Large}
% \subsectionfont{\scshape\normalsize} 
% \subsubsectionfont{\upshape\large}

@@

\newcommand{\sigmapre}{\sigma_\mathrm{pre}}
\newcommand{\sigmapost}{\sigma_\mathrm{post}}

\newcommand{\normalpre}{\mathcal{N}(\mu, \sigmapre)}
\newcommand{\normalpost}{\mathcal{N}(\mu, \sigmapost)}

\newcommand{\ordmeanpre}{\langle o_{\mathrm{pre}}}
\newcommand{\ordmeanpost}{\langle o_{\mathrm{post}}}

@@

Simple consensus can look like group polarization when (1) the mean latent opinion
does not change; (2) both more and less extreme opinions move toward this 
unchanged latent mean (i.e., opinion variance shrinks due to consensus); 
and (3) an ordinal scale is used that bins all extreme
latent opinions into the same extreme bin even though those means have changed,
but opinion bins that are less extreme or of opposite polarity more accurately
reflect the underlying change in latent opinion distribution. This process is
illustrated in Figure~\ref{fig:consensusDistroIllustration}, where we show
a toy example of what can occur when the pre-consensus
opinions (upper left plot)

\begin{table}[h]
  \begin{tabular}{cp{5.0in}} \toprule
   Symbol & Description  \\ \midrule  
   $t \in \{\mathrm{pre}, \mathrm{post}\}$ & Time, either \emph{pre}- or
                                           \emph{post}-discussion \\
   $\mu$  &   Latent mean, assumed constant for simple consensus \\ 
   $\sigma_t$ & Latent standard deviation at time $t$ \\
   $p(o;\mu,\sigma_t)$ & Normal latent distribution of opinions with mean
                         $\mu$ and variance $\sigma_t^2$ \\
   $k \in \{0,\ldots,K\} $ & Threshold index ($k$) between 0 and total number of
         ordinal bins, $K$ \\
   $b_k$ & Opinion bin $k$; note there is no $k=0$ bin. \\
   $\theta_{k-1},~\theta_k$  & Lower, upper threshold value for binning latent opinions into
     bin $b_k$ \\
   $p(o \in b_k; \mu, \sigma_t)$ & Ordinal density of binned opinions in bin $b_k$ 
            (Equation~\ref{eq:binFrequency}) \\
   $\langle o_t \rangle$  &   Observed mean opinion calculated from ordinal data at time $t$ \\
   \bottomrule
  \end{tabular} 
\end{table}

@@

This distributional view models the measurement of $N \to \infty$ participant opinions,
which results in a histogram of frequency of opinions
in each bin, i.e., $o_{i,t} = b_k$. The frequency of responses in each bin is 
the integral over the continuous normal probability density function from 
one bin threshold to another. This transforms the probability density 
function from $p(o_{i,t};~\mu_t, \sigma_t)$ to the probability
of observing a reported opinion of each bin value. 
The probability of observing an opinion in bin
$b_k$ is calculated by integrating the normal probability density 
function over the range of $\theta_{k-1}$ to $\theta_{k}$. 
Formally, we write the 
probability of observing an opinion in bin $k$ at time $t$ as
\begin{equation}
\begin{aligned}
  p(o \in b_k; \mu_t, \sigma_t, \theta) 
    &= \int_{\theta_{k-1}}^{\theta_k} p(o;\mu_t, \sigma_t) do \\
    &= \Phi \left( \frac{o - \mu_t}{\sigma_t} \right)\Big|_{o=\theta_{k-1}}^{\theta_k} \\
    &= \Phi \left( \frac{\theta_k - \mu_t}{\sigma_t} \right) - 
       \Phi \left( \frac{\theta_{k-1} - \mu_t}{\sigma_t} \right)
\end{aligned}
\label{eq:binFrequency}
\end{equation}
\noindent
where $\Phi(\frac{o - \mu}{\sigma})$ is the normalized 
normal cumulative distribution function over opinions $o$, shifted by an amount $\mu$ with
standard deviation $\sigma$.

@@

We have shown in one toy example that group polarization can appear to occur
when ordinal measures are used, without accounting for the binning process; now
we show how to .

We develop two measures for how plausible it is that a reported positive detection
of group polarization is indeed a false positive, building our confidence 
in the plausibility at each step. In this way, we take a modern Bayesian approach
to scientific evidence that typically does not dichotomize ``true'' and ``false''
into absolutes, but evaluates degrees of belief instead.

The first measure we use for whether a reported detection of group polarization
is plausibly false is whether we can find a combination of latent simple
consensus parameters $\mu$, $\sigmapre$, and $\sigmapost$ representing latent
distributions that, when binned (Equation~\ref{eq:binFrequency} and
Figure~\ref{fig:consensusDistroIllustration}), result in ordinal distributions
that have the same difference in means reported in the literature that the authors
took to mean group polarization had occurred. Note that these calculations are
done for the case where effectively we have $N \to \infty$; next we turn to
the question of whether data generated from these simple consensus distributions
generate a ``significant'' group polarization result.

The second measure of how plausible it is that a group polarization is false
is whether ordinal data generated by binning draws from the latent distributions
is found to be significantly different according to Welch's $t$-test.

\subsubsection{Identify null-model parameters that yield plausibly-false positives}

To find simple consensus parameters $\mu$, $\sigmapre$, and $\sigmapost$ that
generate ordinal distributions showing group polarization, we used the above
framework in an optimization routine that takes an initial guess for these latent
parameters and hillclimbs to find $\sigmapre$ and $\sigmapost$ that most accurately
generate ordinal data that appears to have had group polarization occur.

@@

After identifying null-model parameters that yield plausibly-false positives 
in the $N \to \infty$ case, we generated binned samples from normal opinion 
distributions $\normalpre$ and $\normalpost$. Specifically, for each experimental
condition that passed the initial $N \to \infty$ test, we generated 1000 
pre- and post-deliberation ordinal datasets.

We generated these datasets by...

@@

\mt{Is this necessary or can we just reference Liddell \& Kruschke 2018 who show
that ordered-probit models accurately identify latent parameters when normal-distribution
models do not?}
